My Title

### 2022-08-22 misc/collapse cheat sheet/collapse_cheat_sheet_sweave.Rnw
First complete latex cheatsheet.

%% Sets page size and margins
\usepackage[landscape,top=0.8cm,bottom=0.5cm,left=0.5cm,right=0.5cm,marginparwidth=1.75cm]{geometry}

@@

% https://tex.stackexchange.com/questions/36030/how-to-make-a-single-word-look-as-some-code
%\definecolor{light-gray}{gray}{0.95}
%\newcommand{\code}[1]{\colorbox{light-gray}{\texttt{#1}}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\hrrule}{\vspace{2mm}\textcolor{gray}{\hrulefill}\vspace{-2.5mm}} % darkgray
\newcommand{\itxt}[1]{\textcolor{gray}{\textbf{#1}}}

@@

<<setup, echo=FALSE, message=FALSE, warning=FALSE>>=
knitr::opts_chunk$set(warning = FALSE, message = FALSE, cache = TRUE, size = "scriptsize")
library(collapse)
library(magrittr)
library(tibble)
library(microbenchmark)
options(width=80)

@@

{
     {\fontsize{22}{30}\selectfont \textcolor{Gray}{Advanced and Fast Data Transformation with \textit{collapse}}}{\Huge\ : : CHEAT SHEET} %\\%\small{by Sebastian Krantz} %
     \vspace{2mm}
}

@@

\section{\textcolor{WildStrawberry}{Introduction}} % Basics /
%\colorbox{gray}{
\textbf{\textit{collapse}} is a C/C++ based package supporting advanced (grouped, weighted, time series, panel data and recursive) statistical operations in R, with very efficient low-level vectorizations across both groups and columns. \\ [0.8em]

It also offers a flexible, class-agnostic, approach to data transformation in R: handling matrix and data frame based objects in a uniform, attribute preserving, way, and ensuring seamless compatibility with \textit{dplyr} / (grouped) \textit{tibble}, \textit{data.table}, \textit{xts}, \textit{sf} and \textit{plm} classes for panel data ('pseries', 'pdata.frame').  \\ [0.8em]

\textit{collapse} provides full control to the user for statistical programming - with several ways to reach the same outcome and rich optimization possibilities. The default is \code{na.rm = TRUE}, and implemented at low cost at the algorithm level.  \\ [0.8em]

Calling \code{help("collapse-documentation")} brings up a detailed documentation, which is also available \href{https://sebkrantz.github.io/collapse/reference/index.html}{online}. See also the \href{https://fastverse.github.io/fastverse/}{\textit{fastverse}} package/project for a recommended set of complimentary packages and easy package management.
%}

@@

% \vfill\null
\columnbreak
% \hrrule
\section{Fast Statistical Functions}
%\vspace{1mm}
\itxt{Fast functions to perform column–wise grouped and weighted computations on matrix-like objects}
\newline

@@

<<echo=FALSE, include=FALSE>>=
iris <- iris2
@
\vspace{-1mm}

@@

\textsl{collapse} optimizes grouping using both factors / 'qG' objects and 'GRP' objects. 'GRP' objects contain most information and are thus most efficient for complex computations.

<<>>=
X <- iris[1:4]; v <- as.character(iris$Species)
f <- qF(v, na.exclude = FALSE) # Adds 'na.included' class: no NA checks
gv <- group(v) # 'qG' object: first appearance order, with 'na.included'
microbenchmark(fmode(X, v), fmode(X, f), fmode(X, gv), fmode(X, g))
@
\vspace{-2mm}

@@

\section{Fast Data Manipulation}
\vspace{-1.4mm}
\itxt{Efficient implementations of data manipulation functions} % \newline
\setstretch{1.5}
\code{fselect[<-]()} - select/replace columns\\
\code{fsubset()} - subset data (rows and columns)\\
\code{ss()} - fast alternative to \code{[}, particularly for data frames\\
\code{[row|col]order[v]()} - reorder (sort) rows and columns\\
\code{fmutate(), fsummarise()} - dplyr-like, incl. \code{across()} feature\\
\code{[f|set]transform[v][<-]()} - transform cols (by reference)\\
\code{fcompute[v]()} - compute new cols dropping existing ones\\
\code{[f|set]rename()} - rename (any object with 'names' attribute)\\
\code{[set]relabel()} - assign/change variable labels ('label' attr.)\\
\code{get\_vars[<-]()} - select/replace columns (standard eval.)\\ [0.5em]
\setstretch{1}
\code{[num|cat|char|fact|logi|date]\_vars[<-]()} - select/\\ \qquad replace columns by data type or retrieve names/indices\\ [0.5em]
\code{add\_vars[<-]()} - add or column-bind columns \newline

\textbf{Examples}
<<>>=
mtcars |> fsubset(mpg > fnth(mpg, 0.95), disp:wt, cylinders = cyl)
mtcars |> colorder(cyl, vs, am, pos = 'after') |> head(2)
i <- base::invisible # These are equivalent, the second option is faster:
mtcars |> fgroup_by(cyl, vs, am) |> fmutate(sum_mpg = fsum(mpg)) |> i()
mtcars |> fmutate(sum_mpg = fsum(mpg, list(cyl, vs, am), TRA = 1)) |> i()
# These are also equivalent (weighted means), again the second is faster
mtcars |> fgroup_by(cyl) |> fmutate(across(disp:drat, fmean, wt)) |> i()
mtcars |> ftransformv(disp:drat, fmean, cyl, wt, 1, apply = FALSE) |> i()
# ftransform()/fcompute() support list input and ignore attached groupings
mtcars %>% fgroup_by(cyl) %>% ftransform(fselect(., hp:qsec) %>%
           fmedian(TRA = 1) %>% fungroup() %>% fsum(TRA = "/")) |> i()
# Again a faster equivalent: note the use of 'set' to avoid a deep copy
mtcars %>% ftransform(fselect(., hp:qsec) %>% fmedian(cyl, TRA = 1) %>%
                      fsum(TRA = "/", set = TRUE)) %>% i()
# Aggregation: weighted standard deviations
mtcars |> fgroup_by(vs) |> fsummarise(across(disp:drat, fsd, w = wt))
# Grouped linear models: .apply = FALSE applies functions to DF subset
qTBL(mtcars) |> fgroup_by(vs) |> fsummarise(across(disp:drat,
     function(x) list(models = list(lm(disp ~., x))), .apply = FALSE))
# Adding some columns. Use ftransform<- to also replace existing ones
add_vars(iris) <- num_vars(iris) |> fsum(TRA = '%') |> add_stub("perc_")
@
<<echo=FALSE, include=FALSE>>=
iris <- iris2
@
\vspace{-2mm}

@@

\newpage

@@

% \hrrule
\section{Advanced Transformations}
\itxt{Common transformations (in econometrics)} \newline

Scaling, Centering and Averaging\\
\code{fscale(x, g = NULL, w = NULL, na.rm = TRUE,\\
\hphantom{fscaleb}mean = 0, sd = 1, ...)}\\
\code{fwithin(x, g = NULL, w = NULL, na.rm = TRUE,
\hphantom{fwithinb}mean = 0, theta = 1, ...)}\\
\code{fbetween(x, g = NULL, w = NULL, na.rm = TRUE,\\
\hphantom{fbetweenb}fill = FALSE, ...)} \\ [0.5em]

@@

<<>>=
# Grouped scaling
iris |> fgroup_by(Species) |> fscale() |> head(2)
STD(iris, ~ Species, stub = FALSE) |> invisible() # Same thing: faster
# Grouped and weighted scaling. Operators support formulas and keep ids
STD(mtcars, mpg + carb ~ cyl, w = ~ wt) |> head(2)
# Much shorter than fsubset(mpg > fmean(mpg, cyl, TRA = 1))
mtcars |> fsubset(mpg > B(mpg, cyl)) |> head(2)
# Regression with cyl fixed effects - a la Mundlak (1978)
lm(mpg ~ carb + B(carb, cyl), data = mtcars) |> coef()
# Fast grouped (vs) bivariate regression slopes: mpg ~ carb
mtcars |> fgroup_by(vs) |> fmutate(dm_carb = W(carb)) |>
  fsummarise(beta = fsum(mpg, dm_carb) %/=% fsum(dm_carb^2))
# Residuals from regressing on 'Petal' vars and 'Species' FE
fhdwithin(iris[1:2], iris[3:5]) |> head(2)
# Detrending with country-level cubic polynomials
HDW(wlddev, PCGDP + LIFEEX + POP ~ iso3c * poly(year, 3)) |> head(2)
# Note: higher-dimensional centering and polynomials requires fixest
@
 % \emph{Note}: for higher-dimensional centerig / averaging, \emph{collapse} imports \emph{fixest}'s C++ demeaning algorithm, if available.
\vspace{-2mm}

\hrrule
\section{Linear Models}
Fast (barebones) linear model fitting with 6 different solvers\\
\code{flm(y, X, w = NULL, add.icpt = FALSE, method = "lm")}\\ [0.5em]
Fast $R^2$-based F-test of exclusion restrictions for lm's (with FE)\\
\code{fFtest(y, exc, X = NULL, w = NULL, ...)}\\ [0.5em]
Both functions also have formula interfaces:
<<>>=
flm(cbind(mpg, disp) ~ hp + carb, weights = wt, mtcars)
# Test the exclusion of cyl-dummies and hp.
fFtest(mpg ~ qF(cyl) + hp | carb + qF(am), weights = wt, mtcars)
@

<<echo=FALSE>>=
vlabels(wlddev) <- NULL
options(width = 65)
@

% \hrrule
\section{Time Series and Panel Series}
\itxt{Fast and flexible indexed series and data frames}\newline

Turn DF into an 'indexed\_frame' using id and/or time vars\\
\code{data\_ix = findex\_by(data, id1, ..., time)}\\ [0.5em]
\code{data\_ix\$indexed\_series} - contains 'indexed\_series' \\ [0.5em]
\code{index\_df = findex(data\_ix)} - retrieve 'index\_df': DF of id's \\ [0.5em]
\code{with(data\_ix, findex(indexed\_series))} - deep indexation: 'index\_df' can always be retrieved from 'indexed\_series'\\ [0.5em]
\code{is\_irregular(data\_ix)} - check for irregularity in ts/panel\\ [0.5em]
\code{data = unindex(data\_ix)} - remove index\\ [0.5em]
\code{reindex(data, index = index\_df)} - reindexing\\ [0.5em] % (e.g. after \code{readRDS})\\
'indexed\_series' can be 1-or-2D atomic objects. All vectors / time series / matrices can be indexed: \\
\code{reindex(vec/mat, index = vec/index\_df)}\\ [0.5em]
\textbf{Example} \\
% , drop.index.levels = "none"
<<cache=FALSE>>=
wldi <- wlddev |> findex_by(iso3c, year) # Balanced: 216 countries
fsubset(wldi, 1:2, iso3c, year, PCGDP:POP)
# Index stats: [N. ids] | [N. periods (tot.N. periods: (max-min)/GCD)]
LIFEEXi = wldi$LIFEEX # Indexed series
str(LIFEEXi, strict.width = "cut")
LIFEEXi[1:7] # Subsetting indexed series
c(is_irregular(LIFEEXi), is_irregular(LIFEEXi[-5])) # Irregular?
@

\vspace{2mm}
\itxt{Fast functions to perform time-based computations on (irregular) time series and (unbalanced) panel data}\newline

Lags/Leads, Differences, Growth Rates and Cumulative Sums\\
\code{flag(x, n = 1, g = NULL, t = NULL, fill = NA, ...)}\\
\code{fdiff(x, n = 1, diff = 1, ..., log = FALSE, rho = 1)}\\
\code{fgrowth(x, n = 1, diff = 1,, scale = 100, power = 1)}\\
\code{fcumsum(x, g, o, na.rm = TRUE, fill = FALSE)}\newline

Operators:   \code{L(), F(), D(), Dlog(), G()} \newline

\textbf{Example: Computing Growth Rates} \\
<<echo=FALSE, include=FALSE, cache=FALSE>>=
options(width = 70)
wldi = reindex(wldi)
@

<<cache=FALSE>>=
# Ad-hoc use: note that G() supports formulas which fgrowth() doesn't
fgrowth(AirPassengers) |> head()
G(wlddev, c(1, 10), by = PCGDP ~ iso3c, t = ~ year) |> ss(11:12)
wlddev |> fgroup_by(iso3c) |> fselect(iso3c, year, PCGDP, LIFEEX) |>
  fmutate(PCGDP_growth = fgrowth(PCGDP, t = year)) |> head(2)
settransform(wlddev, PCGDP_growth = G(PCGDP, g = iso3c, t = year))
# Note: can omit t -> requires consecutive observations and groups
# Usage with indexed series / frames:
G(wldi) |> head(2) # default: compute growth of num_vars(), keep ids
settransform(wldi, PCGDP_growth = fgrowth(PCGDP))
lm(G(PCGDP) ~ L(G(LIFEEX), 0:3), wldi) |> summary() |> coef() |> round(3)
@

% lm(G(PCGDP) ~ L(G(LIFEEX), 0:3), wldi) |> summary() |> coef() |> round(3)
% flag(vec/mat/DF, n = -1:3, g = id, t = time)
% L(DF, n = -1:2, by = ~ id1 + id2, t = ~ time)
% Use with indexed data or grouped data:
% data_ix |> L(); data_ix |> fmutate(L_var = L(var))
% lm(G(var1) ~ L(G(var2), 0:2), data_ix)
% data |> fgroup_by(id) |> flag(t = time)
% data |> gby(id) |> fmutate(Lv = L(var, t = time))
%
% Cumulative Sums:
% fcumsum(x, g = NULL, o = NULL, fill = FALSE, )
\vspace{2mm}
\code{psacf(), pspacf(), psccf()} - panel series ACF/PACF/CCF\\ [0.5em]
\code{psmat()} - panel data to array conversion/reshaping\\ % \newpage
% \vspace{-5mm}
% <<fig.height=5>>=
% psacf(G(LIFEEXi), main = NA) # ACF of growth rate of life expectancy
% @
%
% <<fig.height=5>>=
% psmat(LIFEEXi) |> plot(colours = TRUE)
% @
%

\hrrule
\section{Summary Statistics}
\code{qsu()} - fast (grouped, weighted, panel-decomposed)\\
\qquad summary statistics for cross-sectional and panel data\\ [0.5em]
\code{qtab()} - faster table() function, incl. weights\\ [0.5em]
\code{descr()} - detailed statistical description of data.frame\\ [0.5em]
\code{varying()} - check variation within groups (panel-ids)\\ [0.5em]
\code{pwcor(), pwcov(), pwnobs()} - pairwise correlations,\\ \qquad covariance and obs. (with P-value and pretty printing)

\hrrule
\section{List-Processing}
%\vspace{1mm}
\itxt{Functions to process (nested) lists (of data objects)}\\ [0.5em]
%\setstretch{1.5}
\code{ldepth()} - level of nesting of list\\ [0.5em]
\code{is\_unlistable()} - is list composed of atomic objects\\ [0.5em]
\code{has\_elem()} - search if list contains certain elements\\ [0.5em]
\code{get\_elem()} - pull out elements from list / subset list\\ [0.5em]
\code{atomic\_elem[<-](), list\_elem[<-]()} - get list with atomic /\\ \qquad sub-list elements, examining only first level of list\\ [0.5em]
\code{reg\_elem(), irreg\_elem()} - get full list tree leading to atomic\\ \qquad ('regular') or non-atomic ('irregular') elements\\ [0.5em]
\code{rsplit()} - efficient (recursive) splitting\\ [0.5em]
\code{t\_list()} - efficient list transpose (transpose lists of lists)\\ [0.5em]
\code{rapply2d()} - recursive apply to lists of data objects\\ [0.5em]
\code{unlist2d()} - recursive row-binding to data.frame\\
%\setstretch{1}

\hrrule
\section{Recode and Replace Values}
\code{recode\_num(), recode\_char()} - recode numeric / character values (+ regex recoding) in matrix-like objects
\code{replace\_NA(), replace\_Inf(), replace\_outliers()} - replace special values   \code{pad()} -  add observations / rows.

\hrrule
\section{(Memory) Efficient Programming}
\itxt{Functions to make R programming more (memory) efficient}\\
\code{anyv}, \code{allv}, \code{allNA}, \code{whichv}, \code{whichNA}, \code{\%==\%},
\code{\%!=\%}, \code{copyv}, \code{setv}, \code{alloc}, \code{missing\_cases}, \code{na\_insert}, \code{na\_rm}, \code{na\_omit}, \code{vlengths}, \code{vtypes}, \code{vgcd}, \code{frange}, \code{fnlevels}, \code{fnrow}, \code{fncol}, \code{fdim}, \code{seq\_row}, \code{seq\_col}, \code{cinv}

<<>>=
x <- na_insert(rnorm(1e7))
microbenchmark(na_rm(x), x[!is.na(x)])
microbenchmark(na.omit(wlddev), na_omit(wlddev, na.attr = TRUE))
microbenchmark(fsubset(wlddev, year == 2010), fsubset(wlddev, year %==% 2010))
# Efficient subassignment by reference
with(wlddev, setv(POP, year %==% 2010, fmax(POP)))
@

\hrrule
\section{Small (Helper) Functions}
\itxt{Functions for (meta-)programming and dealing with attributes}\\
\code{massign}, \code{\%=\%}, \code{.c}, \code{vlabels(<-)}, \code{setLabels}, \code{vclasses}, \code{namlab}, \code{add\_stub}, \code{rm\_stub}, \code{\%!in\%}, \code{ckmatch}, \code{all\_identical}, \code{all\_obj\_equal}, \code{all\_funs}, \code{setDimnames}, \code{setRownames}, \code{setColnames}, \code{unattrib}, \code{setAttrib}, \code{copyAttrib}, \code{copyMostAttrib}, \code{is\_categorical}, \code{is\_date}

\hrrule
\section{Alternative APIs and Global Options}
\itxt{Shorthands for frequently used functions}\\ [0.5em]
\code{fselect -> slt, fsubset -> sbt, fmutate -> mtt, [f/set]transform[v] -> [set]tfm[v], fsummarise -> smr,
across -> acr, fgroup\_by -> gby, findex\_by -> iby, findex -> ix, frename -> rnm, get\_vars -> gv, num\_vars -> nv,
add\_vars -> av} \newline

>>>>

%% Sets page size and margins
\usepackage[landscape,top=0.5cm,bottom=0.6cm,left=0.5cm,right=0.5cm,marginparwidth=1.75cm]{geometry}

@@

% https://tex.stackexchange.com/questions/36030/how-to-make-a-single-word-look-as-some-code
%\definecolor{light-gray}{gray}{0.95}
%\newcommand{\code}[1]{\colorbox{light-gray}{\texttt{#1}}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\hrrule}{\vspace{1mm}\textcolor{gray}{\hrulefill}\vspace{-2.5mm}} % darkgray
\newcommand{\itxt}[1]{\textcolor{gray}{\textbf{#1}}}

@@

<<setup, echo=FALSE, message=FALSE, warning=FALSE>>=
knitr::opts_chunk$set(warning = FALSE, message = FALSE, cache = FALSE, size = "scriptsize")
library(collapse)
library(magrittr)
library(tibble)
library(microbenchmark)
options(width=80)

@@

{
     {\fontsize{22}{30}\selectfont \textcolor{Gray}{Advanced and Fast Data Transformation with \emph{collapse}}}{\Huge\ : : CHEAT SHEET} %\\%\small{by Sebastian Krantz} %
     \vspace{2mm}
}

@@

\section{\textcolor{WildStrawberry}{Introduction}} % Basics /
%\colorbox{gray}{
\textbf{\emph{collapse}} is a C/C++ based package supporting advanced (grouped, weighted, time series, panel data and recursive) statistical operations in R, with very efficient low-level vectorizations across both groups and columns. \\ [0.8em]

It also offers a flexible, class-agnostic, approach to data transformation in R: handling matrix and data frame based objects in a uniform, attribute preserving, way, and ensuring seamless compatibility with \emph{dplyr} / (grouped) \emph{tibble}, \emph{data.table}, \emph{xts}, \emph{sf} and \emph{plm} classes for panel data ('pseries', 'pdata.frame').  \\ [0.8em]

\emph{collapse} provides full control to the user for statistical programming - with several ways to reach the same outcome and rich optimization possibilities. The default is \code{na.rm = TRUE}, and implemented at low cost at the algorithm level.  \\ [0.8em]

Calling \code{help("collapse-documentation")} brings up a detailed documentation, which is also available \href{https://sebkrantz.github.io/collapse/reference/index.html}{online}. See also the \href{https://fastverse.github.io/fastverse/}{\emph{fastverse}} package/project for a recommended set of complimentary packages and easy package management.
%}

@@

% \vfill\null
% \columnbreak
% \hrrule

\section{Fast Statistical Functions}
%\vspace{1mm}
\itxt{Fast functions to perform column–wise grouped and weighted computations on matrix-like objects}
\newline

@@

<<echo=FALSE, include=FALSE>>=
iris <- iris2
@
% \vspace{-1mm}

@@

\emph{collapse} optimizes grouping using both factors / 'qG' objects and 'GRP' objects. 'GRP' objects contain most information and are thus most efficient for complex computations.

<<cache=TRUE>>=
X <- iris[1:4]; v <- as.character(iris$Species)
f <- qF(v, na.exclude = FALSE) # Adds 'na.included' class: no NA checks
gv <- group(v) # 'qG' object: first appearance order, with 'na.included'
microbenchmark(fmode(X, v), fmode(X, f), fmode(X, gv), fmode(X, g))
@
% \vspace{-2mm}

@@

\section{Fast Data Manipulation}
\vspace{-1.4mm}
\itxt{Efficient implementations of data manipulation functions} % \newline
\setstretch{1.5}
\code{fselect[<-]()} - select/replace columns\\
\code{fsubset()} - subset data (rows and columns)\\
\code{ss()} - fast alternative to \code{[}, particularly for data frames\\
\code{[row|col]order[v]()} - reorder (sort) rows and columns\\
\code{fmutate(), fsummarise()} - \emph{dplyr}-like, incl. \code{across()} feature\\
\code{[f|set]transform[v][<-]()} - transform cols (by reference)\\
\code{fcompute[v]()} - compute new cols dropping existing ones\\
\code{[f|set]rename()} - rename (any object with 'names' attribute)\\
\code{[set]relabel()} - assign/change variable labels ('label' attr.)\\
\code{get\_vars[<-]()} - select/replace columns (standard eval.)\\ [0.5em]
\setstretch{1}
\code{[num|cat|char|fact|logi|date]\_vars[<-]()} - select/\\ \qquad replace columns by data type or retrieve names/indices\\ [0.5em]
\code{add\_vars[<-]()} - add or column-bind columns \newline

\textbf{Examples}
<<>>=
mtcars |> fsubset(mpg > fnth(mpg, 0.95), disp:wt, cylinders = cyl)
mtcars |> colorder(cyl, vs, am, pos = 'after') |> head(2)
i <- base::invisible # These are equivalent, the second option is faster:
mtcars |> fgroup_by(cyl, vs, am) |> fmutate(sum_mpg = fsum(mpg)) |> i()
mtcars |> fmutate(sum_mpg = fsum(mpg, list(cyl, vs, am), TRA = 1)) |> i()
# These are also equivalent (weighted means), again the second is faster
mtcars |> fgroup_by(cyl) |> fmutate(across(disp:drat, fmean, wt)) |> i()
mtcars |> ftransformv(disp:drat, fmean, cyl, wt, 1, apply = FALSE) |> i()
# ftransform()/fcompute() support list input and ignore attached groupings
mtcars %>% fgroup_by(cyl) %>% ftransform(fselect(., hp:qsec) %>%
           fmedian(TRA = 1) %>% fungroup() %>% fsum(TRA = "/")) |> i()
# Again a faster equivalent: note the use of 'set' to avoid a deep copy
mtcars %>% ftransform(fselect(., hp:qsec) %>% fmedian(cyl, TRA = 1) %>%
                      fsum(TRA = "/", set = TRUE)) %>% i()
# Aggregation: weighted standard deviations
mtcars |> fgroup_by(vs) |> fsummarise(across(disp:drat, fsd, w = wt))
# Grouped linear models: .apply = FALSE applies functions to DF subset
qTBL(mtcars) |> fgroup_by(vs) |> fsummarise(across(disp:drat,
     function(x) list(models = list(lm(disp ~., x))), .apply = FALSE))
# Adding some columns. Use ftransform<- to also replace existing ones
add_vars(iris) <- num_vars(iris) |> fsum(TRA = '%') |> add_stub("perc_")
@
<<echo=FALSE, include=FALSE>>=
iris <- iris2
@
% \vspace{-2mm}

@@

% \newpage

@@

% \hrrule
\section{Advanced Transformations}
\itxt{Common transformations (in econometrics)} \\ [0.7em]
Scaling, Centering and Averaging\\
\code{fscale(x, g = NULL, w = NULL, na.rm = TRUE,\\
\hphantom{fscaleb}mean = 0, sd = 1, ...)}\\
\code{fwithin(x, g = NULL, w = NULL, na.rm = TRUE,
\hphantom{fwithinb}mean = 0, theta = 1, ...)}\\
\code{fbetween(x, g = NULL, w = NULL, na.rm = TRUE,\\
\hphantom{fbetweenb}fill = FALSE, ...)} \\ [0.5em]

@@

<<>>=
# Grouped scaling
iris |> fgroup_by(Species) |> fscale() |> head(2)
STD(iris, ~ Species, stub = FALSE) |> invisible() # Same thing + faster
# Grouped and weighted scaling. Operators support formulas and keep ids
STD(mtcars, mpg + carb ~ cyl, w = ~ wt) |> head(2)
# Much shorter than fsubset(mpg > fmean(mpg, cyl, TRA = "replace"))
mtcars |> fsubset(mpg > B(mpg, cyl)) |> head(2)
# Regression with cyl fixed effects - a la Mundlak (1978)
lm(mpg ~ carb + B(carb, cyl), data = mtcars) |> coef()
# Fast grouped (vs) bivariate regression slopes: mpg ~ carb
mtcars |> fgroup_by(vs) |> fmutate(dm_carb = W(carb)) |>
  fsummarise(beta = fsum(mpg, dm_carb) %/=% fsum(dm_carb^2))
# Residuals from regressing on 'Petal' vars and 'Species' FE
fhdwithin(iris[1:2], iris[3:5]) |> head(2)
# Detrending with country-level cubic polynomials
HDW(wlddev, PCGDP + LIFEEX + POP ~ iso3c * poly(year, 3)) |> head(2)
# Note: HD centering/prediction and polynomials requires package 'fixest'
@
 % \emph{Note}: for higher-dimensional centerig / averaging, \emph{collapse} imports \emph{fixest}'s C++ demeaning algorithm, if available.
% \vspace{-2mm}

\hrrule
\section{Linear Models}
\vspace{1mm}
Fast (barebones) linear model fitting with 6 different solvers\\
\code{flm(y, X, w = NULL, add.icpt = FALSE, method = "lm")}\\ [0.5em]
Fast $R^2$-based F-test of exclusion restrictions for lm's (with FE)\\
\code{fFtest(y, exc, X = NULL, w = NULL, full.df = TRUE)}\newline

Both functions also have formula interfaces:
<<>>=
flm(cbind(mpg, disp) ~ hp + carb, weights = wt, mtcars)
# Test the exclusion of cyl-dummies and hp.
fFtest(mpg ~ qF(cyl) + hp | carb + qF(am), weights = wt, mtcars)
@

<<echo=FALSE>>=
wlddev2 <- data.table::copy(wlddev)
vlabels(wlddev) <- NULL
options(width = 70)
@

% \hrrule
\section{Time Series and Panel Series}
\itxt{Fast and flexible indexed series and data frames: a modern upgrade of \emph{plm}'s 'pseries' and 'pdata.frame'}\newline

Turn DF into an 'indexed\_frame' using id and/or time vars\\
\code{data\_ix = findex\_by(data, id1, ..., time)}\\ [0.5em]
\code{data\_ix\$indexed\_series} - columns are 'indexed\_series' \\ [0.5em]
\code{index\_df = findex(data\_ix)} - retrieve 'index\_df': DF of ids \\ [0.5em]
\code{index\_df = with(data\_ix, findex(indexed\_series))} - can fetch 'index\_df' from 'indexed\_series' in any caller environment \\ [0.5em]
\code{data = unindex(data\_ix)} - unindex (also 'indexed\_series') \\ [0.5em]
\code{reindex(data, index = index\_df)} - reindex / new pointers\\ [0.5em] % (e.g. after \code{readRDS})\\
'indexed\_series' can be 1-or-2D atomic objects. Vectors / time series / matrices can also be indexed directly using: \\
\code{reindex(vec/mat, index = vec/index\_df)}\\ [0.5em]
\code{is\_irregular()} - irregularity in any index[ed] obj. or time vec\newline

\textbf{Example: Indexing Panel Data} \\
% , drop.index.levels = "none"
<<>>=
wldi <- wlddev |> findex_by(iso3c, year) # Balanced: 216 countries
fsubset(wldi, 1:2, iso3c, year, PCGDP:POP)
# Index stats: [N. ids] | [N. periods (tot.N. periods: (max-min)/GCD)]
LIFEEXi = wldi$LIFEEX # Indexed series
str(LIFEEXi, strict.width = "cut")
LIFEEXi[1:7] # Subsetting indexed series
c(is_irregular(LIFEEXi), is_irregular(LIFEEXi[-5])) # Is irregular?
@
\emph{Note}: 'indexed\_series' and frames are supported via existing 'pseries'/'pdata.frame' methods for time series/panel functions. \newline

% \vspace{2mm}
\itxt{Fast functions to perform time-based computations on (irregular) time series and (unbalanced) panel data}\newline

Lags/Leads, Differences, Growth Rates and Cumulative Sums\\
\code{flag(x, n = 1, g = NULL, t = NULL, fill = NA, ...)}\\
\code{fdiff(x, n = 1, diff = 1, g = NULL, t = NULL,\\
\hphantom{fdiffd}fill = NA, log = FALSE, rho = 1, ...)}\\
\code{fgrowth(x, n = 1, diff = 1, g = NULL, t = NULL, fill\\ \ = NA, logdiff = FALSE, scale = 100, power = 1, ...)}\\
\code{fcumsum(x, g = NULL, o = NULL, na.rm = TRUE,\\
\hphantom{fcumsumd}fill = FALSE, check.o = TRUE, ...)}\\ [0.7em]

Statistical Operators:   \code{L(), F(), D(), Dlog(), G()} \newline

\textbf{Example: Computing Growth Rates} \\
<<echo=FALSE, include=FALSE>>=
# options(width = 70)
wldi = reindex(wldi)
@
<<>>=
# Ad-hoc use: note that G() supports formulas which fgrowth() doesn't
fgrowth(AirPassengers) |> head()
G(wlddev, c(1, 10), by = PCGDP ~ iso3c, t = ~ year) |> ss(11:12)
wlddev |> fgroup_by(iso3c) |> fselect(iso3c, year, PCGDP, LIFEEX) |>
  fmutate(PCGDP_growth = fgrowth(PCGDP, t = year)) |> head(2)
settransform(wlddev, PCGDP_growth = G(PCGDP, g = iso3c, t = year))
# Note: can omit t -> requires consecutive observations and groups
# Usage with indexed series / frames:
@
<<>>=
G(wldi) |> head(2) # default: compute growth of num_vars(), keep ids
settransform(wldi, PCGDP_growth = fgrowth(PCGDP))
lm(G(PCGDP) ~ L(G(LIFEEX), 0:2), wldi) |> summary() |> coef() |> round(3)
@

% lm(G(PCGDP) ~ L(G(LIFEEX), 0:3), wldi) |> summary() |> coef() |> round(3)
% flag(vec/mat/DF, n = -1:3, g = id, t = time)
% L(DF, n = -1:2, by = ~ id1 + id2, t = ~ time)
% Use with indexed data or grouped data:
% data_ix |> L(); data_ix |> fmutate(L_var = L(var))
% lm(G(var1) ~ L(G(var2), 0:2), data_ix)
% data |> fgroup_by(id) |> flag(t = time)
% data |> gby(id) |> fmutate(Lv = L(var, t = time))
%
% Cumulative Sums:
% fcumsum(x, g = NULL, o = NULL, fill = FALSE, )
% \vspace{2mm}
\code{psacf(), pspacf(), psccf()} - panel series ACF/PACF/CCF\\ [0.3em]
\code{psmat()} - panel data to array conversion/reshaping\\ % \newpage
% \vspace{-5mm}
% <<fig.height=5>>=
% psacf(G(LIFEEXi), main = NA) # ACF of growth rate of life expectancy
% @
%
% <<fig.height=5>>=
% psmat(LIFEEXi) |> plot(colours = TRUE)
% @
%
% \vspace{-1mm}
\hrrule
\section{Summary Statistics}
\code{qsu()} - fast (grouped, weighted, panel-decomposed)\\
\qquad summary statistics for cross-sectional and panel data\\
<<>>=
# Panel data statistics: overall, on group-means and group-centered data
qsu(iris, pid = Sepal.Length ~ Species, higher = TRUE)
@
\code{qtab()} - faster \code{table()} function, incl. weights \& custom funs\\ [0.5em]
\code{descr()} - detailed statistical description of data.frame\\ [0.5em]
\code{varying()} - check variation within groups (panel-ids)\\ [0.5em]
\code{pwcor(), pwcov(), pwnobs()} - pairwise correlations,\\ \qquad covariance and obs. (with P-value and pretty printing)
% \vspace{-1mm}

\hrrule
\section{List-Processing}
%\vspace{1mm}
\itxt{Functions to process (nested) lists (of data objects)}\\ [0.5em]
%\setstretch{1.5}
\code{ldepth()} - level of nesting of list\\ [0.5em]
\code{is\_unlistable()} - is list composed of atomic objects\\ [0.5em]
\code{has\_elem()} - search if list contains certain elements\\ [0.5em]
\code{get\_elem()} - pull out elements from list / subset list\\ [0.5em]
\code{atomic\_elem[<-](), list\_elem[<-]()} - get list with atomic /\\ \qquad sub-list elements, examining only first level of list\\ [0.3em]
\code{reg\_elem(), irreg\_elem()} - get full list tree leading to atomic\\ \qquad ('regular') or non-atomic ('irregular') elements\\ [0.5em]
\code{rsplit()} - efficient (recursive) splitting\\ [0.5em]
\code{t\_list()} - efficient list transpose (transpose lists of lists)\\ [0.5em]
\code{rapply2d()} - recursive apply to lists of data objects\\ [0.5em]
\code{unlist2d()} - recursive row-binding to data.frame\\ [0.5em]
%\setstretch{1}

\textbf{Example: Nested Linear Models} \\ %\vspace{-1mm}
<<>>=
(dl <- mtcars |> rsplit(mpg + hp + carb ~ vs + am)) |> str(max.level = 2)
nest_lm <- dl |> rapply2d(lm, formula = mpg ~ .)
(nest_coef <- nest_lm |> rapply2d(summary, classes = "lm") |>
     get_elem("coefficients")) |> str(give.attr = FALSE, strict = "cut")
nest_coef |> unlist2d(c("vs", "am"), row.names = "variable") |> head(2)
@
% \vspace{-2mm}
% \columnbreak

% \hrrule
\section{Recode and Replace Values}
\code{recode\_num(), recode\_char()} - recode numeric / character values (+ regex recoding) in matrix-like objects\\ [0.5em]
\code{replace\_[NA|Inf|outliers]()} - replace special values\\ [0.5em]
\code{pad()} -  add (missing) observations / rows i.e. expand objects
% \vspace{-1mm}

\hrrule
\section{(Memory) Efficient Programming}
\itxt{Functions for (memory) efficient R programming}\\ [0.5em]
